---
title: "Introduction to R and Data Science Workflow: Analyzing Bad Drivers Data"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: readable
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 6)
ft_colors <- c("#FF2700", "#008FD5", "#77787B", "#2FC1D3", "#76B4BD", 
               "#BABABA", "#FFB81C", "#E03A3E")
```

# Welcome to R! 🚗📊

We will learn about R's data science ecosystem through a real-world dataset about driving safety across US states from ["Dear Mona, Which State Has The Worst Drivers?"](https://fivethirtyeight.com/features/which-state-has-the-worst-drivers/)


## Learning Objectives for today


- Work in R/RMarkdown: Running in Console vs Knitting
- Load and explore data using tidyverse
- Visualize data with ggplot2
- Apply the data science workflow: Import → Tidy → Transform → Visualize → Model → Communicate

---

# Part 1: Setting Up Your R Environment 

## Loading Required Packages

In Python, you use `import`. In R, we use `library()`:

```{r load-packages, }
# Install packages if needed (run once)
# install.packages(c("tidyverse", "here", "kableExtra"))

# Load packages (like Python's import)
library(tidyverse)  # Data science toolkit (includes ggplot2, dplyr, readr, etc.)
#library(here)       # File path management
library(kableExtra) # Pretty tables
```

**Python vs R Quick Comparison:**

- Python: `import pandas as pd` → R: `library(tidyverse)`
- Python: `df.head()` → R: `head(df)` or `df |> head()`
- Python: `df.shape` → R: `dim(df)`

---

# Part 2: Data Import and First Look 

```{r import-data}
# Import data directly from FiveThirtyEight's GitHub
# This is like pd.read_csv() in Python
drivers <- read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/bad-drivers/bad-drivers.csv")

# First look at the data
#drivers
```

## Data Exploration

```{r explore-data}
# Dimensions (like df.shape in Python)
dim(drivers)

# Column names and types
glimpse(drivers)

# Summary statistics (like df.describe() in Python)
summary(drivers)
```

## Clean Up Column Names

```{r clean-names}
# Some data wrangling
# objects and object assignment
# <- object assignmnet
# pipe:  |> or %>%
# Rename columns for easier analysis
drivers_clean <- drivers |>
  rename(
    state = State,
    fatal_rate = `Number of drivers involved in fatal collisions per billion miles`,
    pct_speeding = `Percentage Of Drivers Involved In Fatal Collisions Who Were Speeding`,
    pct_alcohol = `Percentage Of Drivers Involved In Fatal Collisions Who Were Alcohol-Impaired`,
    pct_not_distracted = `Percentage Of Drivers Involved In Fatal Collisions Who Were Not Distracted`,
    pct_no_previous = `Percentage Of Drivers Involved In Fatal Collisions Who Had Not Been Involved In Any Previous Accidents`,
    insurance_premium = `Car Insurance Premiums ($)`,
    losses = `Losses incurred by insurance companies for collisions per insured driver ($)`
  )


# Check 
names(drivers_clean)
head(drivers_clean)
```

```{r}
## Some moredata wrangling
## 
## Most dangerous states (by fatal collision rate):
drivers_clean %>%
  select(state,  fatal_rate) # %>%
#  arrange(desc(fatal_rate)) %>%
#  head(5) %>%
#  knitr::kable()
```

---

# Part 3: Data Visualization with ggplot2 

ggplot2 uses a "grammar of graphics" - you build plots layer by layer.

**Basic ggplot2 syntax:**
```r
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

The `<GEOM_FUNCTION>` part tells R what kind of plot you want to make. Some of the most common geoms:

- geom_point() → scatterplots
- geom_bar() → bar plots (counts of categories)
- geom_col() → bar plots when you already have values to plot (not counts)
- geom_line() → line plots
- geom_histogram() → histograms
- geom_boxplot() → boxplots
- geom_density() → smoothed density curves

## Visualization 1: Fatal Collision Rates by State

```{r}
# basic plot
ggplot(data = drivers_clean)
```


```{r plot1-fatal-rates0}
# Basic bar chart
ggplot(data = drivers_clean) +
  geom_col(mapping = aes(x = state, y = fatal_rate))
```



### Improve the plot 




## TO DO LATER: Assignment 5 Plot 1: 

Come up with a neat visualization that shows something interesting about insurance premiums

```{r}
## Add your code here
```

## Your Turn(YT) Distribution of a variable/column (random variable)

```{r your-turn-1, eval=FALSE}
# Create a histogram showing the distribution of insurance premiums
# Fill in the blanks and remove eval=FALSE while knitting

drivers_clean %>%
  ggplot(aes(x = _______)) +
  geom_______(_____) +
  labs(
    title = "Distribution of average annual car insurance Premiums",
    subtitle = "______",
    x = "_______",
    y = "_______"
  ) +
  theme_minimal()
```

## Visualization 2: Explore relationships amongst variables:

**Jiho Shin**: Alcohol and Fatal Collisions


```{r reln-alcohol-fatalities1}
ggplot(drivers_clean, aes(x = pct_alcohol, y = fatal_rate)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = "Alcohol-Impaired Driving vs Fatal Collisions",
    x = "% of Drivers Alcohol-Impaired in Fatal Collisions",
    y = "Fatal Collisions per Billion Miles"
  ) +
  theme_minimal()
```

```{r reln-alcohol-fatalities2}
ggplot(drivers_clean, aes(x = pct_alcohol, y = fatal_rate)) +
  geom_point(color = "firebrick", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(
    title = "Relationship between Fatal Collisions and  Alcohol-Impaired Driving",
    subtitle = "",
    x = "% of Drivers Alcohol-Impaired in Fatal Collisions",
    y = "Fatal Collisions per Billion Miles"
  ) +
  theme_minimal()
```


## Visualization 3: Comparing Causes: Alcohol, Speeding, and Distraction


```{r compare-alchol-speeding-distraction}
## Multiple scatter plot 

## data wrangling
drivers_long <- drivers_clean |>
  select(state, fatal_rate, pct_speeding, pct_alcohol, pct_not_distracted) |>
  mutate(
    pct_distracted = 100 - pct_not_distracted
  ) |>
  select(state, fatal_rate, pct_speeding, pct_alcohol, pct_distracted) |>
  pivot_longer(cols = starts_with("pct"),
               names_to = "cause",
               values_to = "percent")

ggplot(drivers_long, aes(x = percent, y = fatal_rate, color = cause)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c(
    pct_speeding = "steelblue",
    pct_alcohol = "firebrick",
    pct_distracted = "darkgreen"
  ),
  labels = c(
    pct_speeding = "Speeding",
    pct_alcohol = "Alcohol",
    pct_distracted = "Distracted")
  ) +
  labs(
    title = "Comparing Causes of Fatal Collisions",
    subtitle = "As alcohol involvement rises, fatal collisions increase noticeably",
    x = "% of Drivers Involved in Fatal Collisions",
    y = "Fatal Collisions per Billion Miles",
    color = "Cause"
  ) +
  theme_minimal()
```




## Visualization 4: Regional Analysis

```{r plot4-regions}
# Add regions (you could also join with a separate dataset)
drivers_regions <- drivers_clean |>
  mutate(
    region = case_when(
      state %in% c("Maine", "New Hampshire", "Vermont", "Massachusetts", 
                   "Rhode Island", "Connecticut", "New York", "New Jersey", "Pennsylvania") ~ "Northeast",
      state %in% c("Ohio", "Michigan", "Indiana", "Illinois", "Wisconsin", 
                   "Minnesota", "Iowa", "Missouri", "North Dakota", "South Dakota", 
                   "Nebraska", "Kansas") ~ "Midwest",
      state %in% c("Delaware", "Maryland", "District of Columbia", "Virginia", 
                   "West Virginia", "Kentucky", "Tennessee", "North Carolina", 
                   "South Carolina", "Georgia", "Florida", "Alabama", "Mississippi", 
                   "Arkansas", "Louisiana", "Texas", "Oklahoma") ~ "South",
      state %in% c("Montana", "Wyoming", "Colorado", "New Mexico", "Idaho", 
                   "Utah", "Nevada", "Arizona", "Washington", "Oregon", 
                   "California", "Alaska", "Hawaii") ~ "West"
    )
  )

# Box plot  of fatal collisiion rate by region
ggplot(drivers_regions, aes(x = region, y = fatal_rate, fill = region)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(
    title = "Distribution of Fatal Collision Rates by US Region",
    subtitle = "",
    x = "Region",
    y = "Fatal Collision Rate (per billion miles)",
    caption = "Data: FiveThirtyEight"
  ) +
  theme_minimal() +
  theme(legend.position = "none")  # Remove legend since x-axis shows regions
```

## TO DO LATER: Assignment 5 Plot 2: 

How do insurance premiums across regions? Show an interesting plot and don't forget to conclude with your written brief take-away. 

```{r}
## Add your code here
```


# Part 4: Data Transformation and Summary Statistics

## Creating Summary Statistics

```{r summary-stats}
# Summary by region
regional_summary <- drivers_regions |>
  group_by(region) |>
  summarise(
    n_states = n(),
    avg_fatal_rate = mean(fatal_rate, na.rm = TRUE),
    median_fatal_rate = median(fatal_rate, na.rm = TRUE),
    avg_insurance = mean(insurance_premium, na.rm = TRUE),
    avg_speeding = mean(pct_speeding, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(avg_fatal_rate))

# Display as a nice table
regional_summary |>
  kable(
    caption = "Regional Summary of Driving Safety Metrics",
    digits = 2,
    col.names = c("Region", "# States", "Avg Fatal Rate", "Median Fatal Rate", 
                  "Avg Insurance Premium", "Avg % Speeding")
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Find the "Worst" and "Best" Drivers

```{r extreme-states}
# States with highest fatal rates
worst_drivers <- drivers_clean |>
  slice_max(fatal_rate, n = 5) |>
  select(state, fatal_rate, pct_speeding, pct_alcohol, insurance_premium)

# States with lowest fatal rates
best_drivers <- drivers_clean |>
  slice_min(fatal_rate, n = 5) |>
  select(state, fatal_rate, pct_speeding, pct_alcohol, insurance_premium)

print("States with Highest Fatal Collision Rates:")
worst_drivers |>
  kable(digits = 2) |>
  kable_styling(bootstrap_options = c("striped", "hover"))

print("States with Lowest Fatal Collision Rates:")
best_drivers |>
  kable(digits = 2) |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

## Part 5: Advanced plotting

### Choropleth Map of Fatal Collision Rates



### Interactive Map with Leaflet


---

# Part 6: Putting It All Together 

## Key Findings

Based on our analysis:

1. **Fatal collision rates vary significantly** across states, from `r round(min(drivers_clean$fatal_rate), 1)` to `r round(max(drivers_clean$fatal_rate), 1)` per billion miles.

2. **Regional patterns exist**: Some regions show consistently higher or lower rates.

3. **Multiple factors contribute**: Speeding, alcohol use, and distraction all play roles, but their relationships with fatal rates are complex.

4. **Insurance costs reflect risk**: States with higher fatal rates tend to have higher insurance premiums.

## The Data Science Workflow in Action

We've completed a full cycle:

1. **Import**: Loaded data from FiveThirtyEight
2. **Tidy**: Cleaned column names, reshaped data for analysis  
3. **Transform**: Created regional groupings, calculated summaries
4. **Visualize**: Created multiple plot types to explore patterns
5. **Model**: Used trend lines to explore relationships
6. **Communicate**: Organized findings in a clear report

---

# Getting Help

- **AI**
- **R Documentation**: Use `?function_name` (e.g., `?ggplot`)
- **Tidyverse Website**: https://www.tidyverse.org/
- **R for Data Science Book**: https://r4ds.had.co.nz/
- **Stack Overflow**: Tag your questions with [r] and [ggplot2]

**Remember**: R has a great community. When you get stuck, someone has probably asked the same question before!

---
